\Title

### 2019-06-06 manuscript/manuscript.tex
Format a version for the EarthArxiv
Use the Latex article class instead of GJI and insert the self-archiving
text from the publisher.

%\documentclass[extra]{gji}
\documentclass[extra, referee]{gji}

\usepackage[utf8]{inputenc}
\usepackage{timet}
\usepackage{amsmath}
\usepackage{graphicx}

\usepackage{url}
\usepackage[pdftex,colorlinks=true]{hyperref}
\hypersetup{
    allcolors=blue,
}

@@

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, M.E. Gimenez, and L. Uieda]{
    Santiago R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$,
    Mario E. Gimenez$^{1,2}$, and Leonardo Uieda$^3$ \\
    $^1$CONICET, Argentina.~e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Earth Sciences, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

\maketitle

\begin{summary}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear,
exponential, and sinusoidal density functions.
The heavily oscillating density functions are intended only to test the algorithm to its
limits and not to emulate a real world case.
These comparisons allowed us to obtain optimal values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential and
its gradient are 1 and 2.5, respectively.
The density-based discretization algorithm produces no discretizations in the linear
density case, but a delta ratio of 0.1 is needed for the exponential and most sinusoidal
density functions.
These values can be extrapolated to cover most common use cases, which are simpler than
oscillating density profiles.
However, the distance-size and delta ratios can be configured by the user to increase
the accuracy of the results at the expense of computational speed.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000~m, using an exponential density function.
\end{summary}

\begin{keywords}
Numerical modelling, Numerical approximations and analysis, Gravity anomalies
and Earth structure, Satellite gravity
\end{keywords}

@@

\begin{figure}
\centering
\includegraphics[width=0.6\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
    A tesseroid (spherical prism) in a geocentric spherical coordinate system, with a
    computation point $P$ and its local north oriented Cartesian coordinate system.
    After \citet{Uieda2015}.
}
\label{fig:tesseroid}
\end{figure}

@@

\iftwocol{
\begin{equation}
    \begin{split}
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \rho(r') f(r', \phi', \lambda')
        dr' d\phi' d\lambda' \approx& \\
        A
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        W_i^r W_j^\phi W_k^\lambda
        &\rho(r_i) f(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') f(r', \phi', \lambda')
    dr' d\phi' d\lambda' \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

@@

\begin{table}
\caption{
    Description of the tesseroid models used to build spherical shells and characterize
    the accuracy of the numerical integration.
    The outer radius ($R_2$) of every shell model is equal to the mean Earth radius
    (6378.137 km), while the inner radius ($R_1$) is determined by its thickness.
    The horizontal dimensions of the tesseroids and the total number of
    tesseroids in the shell model are given in the latitudinal and longitudinal
    dimensions, respectively.
}
\label{tab:shell-models}
\begin{tabular}{rccccc}
    Thickness & Tesseroid size & Number of tesseroids \\ \hline
    0.1 km  & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    1 km    & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    10 km   & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    100 km  & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    1000 km & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
\end{tabular}
\end{table}

\begin{table}
\caption{
    Description of the computation grids used to characterize the accuracy of the
    numerical integration.
    Grid height is defined above the mean Earth radius.
}
\label{tab:grids}
\begin{tabular}{lccc}
    Name & Grid spacing & Grid region (degrees) & Grid height (km)
    \\ \hline
    Pole      & $0.1^\circ$ &   0E /   1E / 89N / 90N & 0   \\
    Equator   & $0.1^\circ$ &   0E /   1E /  0N / 1N  & 0   \\
    Global    & $ 10^\circ$ & 180W / 180E / 90S / 90N & 0   \\
    Satellite & $ 10^\circ$ & 180W / 180E / 90S / 90N & 260 \\
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=\linewidth]{figures/exponential-densities.pdf}
}{
\includegraphics[width=0.5\linewidth]{figures/exponential-densities.pdf}
}
\caption{
    Exponential density functions assigned to the spherical shell models for
    $\delta$ ratio determination.
    Each density function corresponds to a different value of $b$ on
    Eq.~\ref{eq:density-exp}.
}
\label{fig:exp-densities}
\end{figure}

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=\linewidth]
    {figures/grid-search.pdf}
}{
\includegraphics[width=0.5\linewidth]
    {figures/grid-search.pdf}
}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison between the
    analytical solution and the numerical approximation of the gravitational fields ($V$
    and $g_z$) generated by a spherical shell with an exponential density function
    (Eq.~\ref{eq:density-exp}).
    These comparisons were carried out on the ``Global'' grid (Table~\ref{tab:grids}),
    with the spherical shell models detailed in Table~\ref{tab:shell-models}, and an
    exponential density function with $b = 30$.
    The percentage difference values are obtained as the maximum difference
    between every shell model.
    The points inside the dashed line are the ones that present an error lower than
    0.1\%.
    The $D$ value obtained for the linear density case for each gravitational field is
    also shown ($D_\text{linear}$).
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=\linewidth]{figures/sine-densities.pdf}
}{
\includegraphics[width=0.5\linewidth]{figures/sine-densities.pdf}
}
\caption{
    Sinusoidal density functions assigned to the spherical shells in the $\delta$ ratio
    determination.
    Each density function corresponds to a different value of $b$ on
    Eq.~\ref{eq:density-sine}.
}
\label{fig:sine-densities}
\end{figure}

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=\linewidth]{figures/neuquen-basin-densities.pdf}
}{
\includegraphics[width=0.5\linewidth]{figures/neuquen-basin-densities.pdf}
}
\caption{
    Linear and exponential densities used to compute the gravitational fields generated
    by a tesseroid model of the Neuqu\'en sedimentary basin.
    The height is defined above the mean Earth radius, and its axis is spanned between
    the deepest and the highest point of the basin.
}
\label{fig:neuquen-basin-densities}
\end{figure}

@@

\begin{equation}
    \begin{split}
    V_\text{sine}(r) = \frac{4\pi G}{r} \frac{A}{k^3} \Big[
    & (2 - k^2 R_2^2) \cos(k(R_2 - R)) + 2 k R_2 \sin(k(R_2 - R)) - \\
    & (2 - k^2 R_1^2) \cos(k(R_1 - R)) - 2 k R_1 \sin(k(R_1 - R))
        \Big].
    \end{split}
\end{equation}

>>>>

\documentclass[twocolumn]{article}

\newcommand{\Title}{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\newcommand{\Author}{S.R. Soler, A. Pesce, M.E. Gimenez, L. Uieda}
\newcommand{\AuthorAffil}{
    {\large
        Santiago R. Soler$^{1,2,*}$, Agustina Pesce$^{1,2}$,
        Mario E. Gimenez$^{1,2}$, and Leonardo Uieda$^3$
    }
    \\[0.4cm]
    {\small $^1$ CONICET, Argentina}
    \\
    {\small $^2$ Instituto Geofísico Sismológico Volponi, Universidad Nacional de San Juan, Argentina}
    \\
    {\small $^3$ Department of Earth Sciences, SOEST, University of Hawai'i at M\={a}noa, USA}
    \\
    {\small $^*$ e-mail: santiago.r.soler@gmail.com}
}

\usepackage[left=0.7in,right=0.7in,top=1in,bottom=1in]{geometry}
\setlength{\columnsep}{2\columnsep}
\usepackage[utf8]{inputenc}
\usepackage{timet}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[round]{natbib}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage[pdftex,colorlinks=true]{hyperref}
\hypersetup{
    allcolors=blue,
    pdftitle={\Title},
    pdfauthor={\Author},
}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\lhead{
    \fontsize{9pt}{12pt}\selectfont
    \Author (2019).
    Accepted for publication in \textit{Geophysical Journal International}.
}
\rhead{\fontsize{9pt}{12pt}\selectfont \thepage}
\renewcommand{\headrulewidth}{0pt}

@@

\title{\Title}
\author{\AuthorAffil}
\date{
    \normalsize
    Accepted 2019 June 05. Received 2019 May 10; in original form 2018 December 29
    \\[0.4cm]
    This is a pre-copyedited, author-produced PDF of an article accepted for
    publication in \textit{Geophysical Journal International} following peer review.
    The version of record
    ``\textit{INSERT COMPLETE CITATION INFORMATION HERE}''
    is available online at: \href{https://doi.org/XXXXXXX}{doi.org/XXXXXX}
}
\maketitle

\begin{abstract}
We present a new methodology to compute the gravitational fields generated by
tesseroids (spherical prisms) whose density varies with depth according to
an arbitrary continuous function.
It approximates the gravitational fields through the Gauss-Legendre Quadrature along
with two discretization algorithms that automatically control its accuracy by adaptively
dividing the tesseroid into smaller ones.
The first one is a preexisting two dimensional adaptive discretization algorithm that
reduces the errors due to the distance between the tesseroid and the computation point.
The second is a new density-based discretization algorithm that
decreases the errors introduced by the variation of the density function with depth.
The amount of divisions made by each algorithm is indirectly controlled
by two parameters: the distance-size ratio and the delta ratio.
We have obtained analytical solutions for a spherical shell with radially variable
density and compared them to the results of the numerical model for linear,
exponential, and sinusoidal density functions.
The heavily oscillating density functions are intended only to test the algorithm to its
limits and not to emulate a real world case.
These comparisons allowed us to obtain optimal values for the distance-size and
delta ratios that yield an accuracy of 0.1\% of the analytical solutions.
The resulting optimal values of distance-size ratio for the gravitational potential and
its gradient are 1 and 2.5, respectively.
The density-based discretization algorithm produces no discretizations in the linear
density case, but a delta ratio of 0.1 is needed for the exponential and most sinusoidal
density functions.
These values can be extrapolated to cover most common use cases, which are simpler than
oscillating density profiles.
However, the distance-size and delta ratios can be configured by the user to increase
the accuracy of the results at the expense of computational speed.
Lastly, we apply this new methodology to model the Neuqu\'en Basin, a foreland basin in
Argentina with a maximum depth of over 5000~m, using an exponential density function.
\\[0.5cm]
\textbf{Keywords:}
Numerical modelling, Numerical approximations and analysis, Gravity anomalies
and Earth structure, Satellite gravity
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
    A tesseroid (spherical prism) in a geocentric spherical coordinate system, with a
    computation point $P$ and its local north oriented Cartesian coordinate system.
    After \citet{Uieda2015}.
}
\label{fig:tesseroid}
\end{figure}

@@

%\iftwocol{
\begin{equation}
    \begin{split}
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \rho(r') f(r', \phi', \lambda')
        dr' d\phi' d\lambda' \approx& \\
        A
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        W_i^r W_j^\phi W_k^\lambda
        &\rho(r_i) f(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
%}{
%\begin{equation}
    %\int\limits_{\lambda_1}^{\lambda_2}
    %\int\limits_{\phi_1}^{\phi_2}
    %\int\limits_{r_1}^{r_2}
    %\rho(r') f(r', \phi', \lambda')
    %dr' d\phi' d\lambda' \approx
    %A
    %\sum\limits_{i=1}^{N^r}
    %\sum\limits_{j=1}^{N^\phi}
    %\sum\limits_{k=1}^{N^\lambda}
    %W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
%\label{eq:glq-var-dens}
%\end{equation}
%}

@@

\begin{table}
\caption{
    Description of the tesseroid models used to build spherical shells and characterize
    the accuracy of the numerical integration.
    The outer radius ($R_2$) of every shell model is equal to the mean Earth radius
    (6378.137 km), while the inner radius ($R_1$) is determined by its thickness.
    The horizontal dimensions of the tesseroids and the total number of
    tesseroids in the shell model are given in the latitudinal and longitudinal
    dimensions, respectively.
    \newline
}
\label{tab:shell-models}
\centering
\begin{tabular}{rccccc}
    Thickness & Tesseroid size & Number of tesseroids \\ \hline
    0.1 km  & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    1 km    & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    10 km   & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    100 km  & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    1000 km & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
\end{tabular}
\end{table}

\begin{table*}
\caption{
    Description of the computation grids used to characterize the accuracy of the
    numerical integration.
    Grid height is defined above the mean Earth radius.
    \newline
}
\label{tab:grids}
\centering
\begin{tabular}{lccc}
    Name & Grid spacing & Grid region (degrees) & Grid height (km)
    \\ \hline
    Pole      & $0.1^\circ$ &   0E /   1E / 89N / 90N & 0   \\
    Equator   & $0.1^\circ$ &   0E /   1E /  0N / 1N  & 0   \\
    Global    & $ 10^\circ$ & 180W / 180E / 90S / 90N & 0   \\
    Satellite & $ 10^\circ$ & 180W / 180E / 90S / 90N & 260 \\
\end{tabular}
\end{table*}

@@

\begin{figure}
\centering
%\iftwocol{
\includegraphics[width=\linewidth]{figures/exponential-densities.pdf}
%}{
%\includegraphics[width=0.5\linewidth]{figures/exponential-densities.pdf}
%}
\caption{
    Exponential density functions assigned to the spherical shell models for
    $\delta$ ratio determination.
    Each density function corresponds to a different value of $b$ on
    Eq.~\ref{eq:density-exp}.
}
\label{fig:exp-densities}
\end{figure}

@@

\begin{figure}
\centering
%\iftwocol{
\includegraphics[width=\linewidth]
    {figures/grid-search.pdf}
%}{
%\includegraphics[width=0.5\linewidth]
    %{figures/grid-search.pdf}
%}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison between the
    analytical solution and the numerical approximation of the gravitational fields ($V$
    and $g_z$) generated by a spherical shell with an exponential density function
    (Eq.~\ref{eq:density-exp}).
    These comparisons were carried out on the ``Global'' grid (Table~\ref{tab:grids}),
    with the spherical shell models detailed in Table~\ref{tab:shell-models}, and an
    exponential density function with $b = 30$.
    The percentage difference values are obtained as the maximum difference
    between every shell model.
    The points inside the dashed line are the ones that present an error lower than
    0.1\%.
    The $D$ value obtained for the linear density case for each gravitational field is
    also shown ($D_\text{linear}$).
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{figure}
\centering
%\iftwocol{
\includegraphics[width=\linewidth]{figures/sine-densities.pdf}
%}{
%\includegraphics[width=0.5\linewidth]{figures/sine-densities.pdf}
%}
\caption{
    Sinusoidal density functions assigned to the spherical shells in the $\delta$ ratio
    determination.
    Each density function corresponds to a different value of $b$ on
    Eq.~\ref{eq:density-sine}.
}
\label{fig:sine-densities}
\end{figure}

@@

\begin{figure}
\centering
%\iftwocol{
\includegraphics[width=\linewidth]{figures/neuquen-basin-densities.pdf}
%}{
%\includegraphics[width=0.5\linewidth]{figures/neuquen-basin-densities.pdf}
%}
\caption{
    Linear and exponential densities used to compute the gravitational fields generated
    by a tesseroid model of the Neuqu\'en sedimentary basin.
    The height is defined above the mean Earth radius, and its axis is spanned between
    the deepest and the highest point of the basin.
}
\label{fig:neuquen-basin-densities}
\end{figure}

@@

\begin{equation}
    \begin{split}
        V_\text{sine}(r) = \frac{4\pi G}{r} \frac{A}{k^3} \Big[
    & (2 - k^2 R_2^2) \cos(k(R_2 - R)) + \\
    & 2 k R_2 \sin(k(R_2 - R)) - \\
    & (2 - k^2 R_1^2) \cos(k(R_1 - R)) - \\
    & 2 k R_1 \sin(k(R_1 - R))
    \Big].
    \end{split}
\end{equation}
